---
title: "Code_Thesis"
output: notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Notations & important formulas

- Matrix Ds : include the binary mutational profiles of ns different single cells on the set of m somatic mutations
- Expected genotype matrix Gs : binary matrix. Row = single cell of the sth experiment. Column = a mutation or an aggregate mutational event.
- $$α$$ (alpha) : False positive rates
- $$β$$ (beta) : False negative rates



Weighted likelihood :

$$
P(\{\textbf{D}\}_{s=1}^y | \{\textbf{G}_s\}_{s=1}^y , \{\alpha_s\}_{s=1}^y,\{\beta_s\}_{s=1}^y) = \prod_{s=1}^y P(\textbf{D}_s | \textbf{G}_s, \alpha_s, \beta_s)^{w_s}
$$
We employ the following formula to compute the previous one : 

$$ 
P(\textbf{D}_s | \textbf{G}_s, \alpha_s, \beta_s) = \prod_{i=1}^n\prod_{j=1}^m P(d_{i,j}^s | g_{i,j}^s, \alpha_s, \beta_s)
$$
where we have :

$$
P(d_{i,j}^s | g_{i,j}^s, \alpha_s, \beta_s) = \left\{
  \begin{array}{l}
    \alpha_s, & if &  d_{i,j}^s=1 & and &g_{i,j}^s=0, \\
    1- \alpha_s, &  if & d_{i,j}^s=1 & and & g_{i,j}^s=1,\\
    \beta_s, &  if &  d_{i,j}^s=0 & and & g_{i,j}^s=1,\\
    1- \beta_s,  &  if & d_{i,j}^s=0 & and & g_{i,j}^s=0,\\
    1,   & if  & d_{i,j}^s=NA, \\
  \end{array}
\right.

$$

## Thesis context

A likelihood function is used in the LACE method.The approach consist on maximizing the weighted likelihood function.
One of the problems that can occur with a large amount of data is noise. In this case we can have error rates (alpha and beta as defined before). We can also have no value (represented as NA).
These errors and especially the NA can have a negative impact on the calculation (computing).
The goal here is to reduce this impact and obtain the best result for the likelihood.

We define the likelihood for one matrix with this formula : 

$$
L_i = \alpha^{FP}(1-\alpha)^{n-FP}\beta^{FN}(1-\beta)^{n-FN}
$$
To have the overall likelihood we have the following formula : 

$$
L = \sum_{i=1}^n L_i
$$
If we want the weighted likelihood we just have to add the weight in the formula : 

$$
L = \sum_{i=1}^n L_i*w_i
$$
## Practical part

################Initialization################
```{r Initialisation}

NM=20                   #Number of matrices
i=60                    #Lines in each matrix
j=4                     #Columns in each matrix
list_likelihood = c()   #Contains all likelihood values for all matrices
list_NA = c()           #Contains the number of NA in each matrix
list_weight = c()       #Contains all weight values for all matrices
list_alpha = c()        #Contains all alpha values for all matrices
list_beta = c()         #Contains all beta values for all matrices
list_FP = c()           #Contains the number of FP in each matrix
list_FN = c()           #Contains the number of FN in each matrix
df1 = data.frame(matrix(ncol = 0, nrow = 0))
order_df = data.frame(matrix(ncol = 0, nrow = 0))
```

################Generation of an example################

``` {r Generation}

#This function allows to create the different lists needed to generate different examples to test the idea

lists <- function(NM,i,j){ #NM = Number of matrices used to calculate the global likelihood
  list_alpha <<- c(round(runif(NM,0,0.5), digit = 2))
  list_beta <<- c(round(runif(NM,0,0.5), digit = 2))
  #list_NA <<- c(sample(0:((i*j)/3),NM, replace=F))
  #list_FP <<- c(sample(0:((i*j)/3),NM, replace=F))
  #list_FN <<- c(sample(0:((i*j)/3),NM, replace=F))
  list_NA <<- c(round(runif(NM,0,(i*j)/3), digit = 0))
  list_FP <<- c(round(runif(NM,0,(i*j)/3), digit = 0))
  list_FN <<- c(round(runif(NM,0,(i*j)/3), digit = 0))
}

lists(NM,i,j)
```

################Likelihood################

``` {r Likelihood}

#This function calculates the likelihood value for each of the matrices. The formula is the one we mentioned before.

likelihood <-function(NM,i,j,alpha,beta,FP, FN){
  list_likelihood <<- c()
  for (i in 1:NM){
    n=i*j
    L <- (alpha[i]^FP[i])*((1-alpha[i])^(n-FP[i]))*(beta[i]^FN[i])*((1-beta[i])^(n-FN[i]))
      list_likelihood <<- append(list_likelihood, L)
  }
}

loglikelihood <-function(NM,i,j,alpha,beta,FP, FN){ #List_NA,alpha,beta,
  list_likelihood <<- c()
  for (i in 1:NM){
    n=i*j
    L <- -log((alpha[i]^FP[i])*((1-alpha[i])^(n-FP[i]))*(beta[i]^FN[i])*((1-beta[i])^(n-FN[i])))
      list_likelihood <<- append(list_likelihood, L)
  }
}

#likelihood(NM,i,j,list_alpha, list_beta, list_FP, list_FN)
loglikelihood(NM,i,j,list_alpha, list_beta, list_FP, list_FN)
while((Inf %in% list_likelihood)== TRUE) {
  lists(NM,i,j)
  loglikelihood(NM,i,j,list_alpha, list_beta, list_FP, list_FN)
}
list_likelihood
```

################Ordering the list of NA & creation of the matrix of values################

``` {r Odered NA & matrix of values}

#This part allows to order the list of NA to have the matrices with the smallest number of NA first, and the one with the most NA last.
#We then create a table containing the likelihood values and the NA number for each matrix

NA_order <- function(){
  df1 <<- data.frame("Likelihood" = list_likelihood, "NumberNA" = list_NA)
  order_df <<- df1[order(df1$"NumberNA"),]
}

NA_order()
df1
order_df
```

################Definition of the global weight################

``` {r Global weight}

#This function calculates the overall starting value for the weight. 
#Remember that the sum of the weights must be equal to 1.
#We therefore take this value and divide it by the number of matrices we have.

global_weights <- function(NM) {
  gw <- 1/ NM
  list_weight <<- c()
  for (i in 1:NM) {
    list_weight[i] <<- gw
  }
}

global_weights(NM)
list_weight
```

################Definition of ordered weights################

``` {r Ordered weights}

#Now, the reasoning is that we rely on the number of NA present in each matrix.
#We modify the weight values to obtain a progressive list of values according to the previous criterion
#For example, if we have 5 matrices instead of having the values : (0.2,0.2,0.2,0.2,0.2)
#we will have : (0.4,0.3,0.2,0.1,0)
#We take into account if the vector has an even or odd length 

weights <- function(NM){
  mid <<-ceiling(NM/2)
  if (NM%%2 == 0){ #if the number of matrices is odd or even
    lim <<- mid+1
  } else {lim <<- mid }
  for (p in 1:lim) {
    if (p == mid || p==lim) {
      list_weight[p] <<- list_weight[p]
    } else {
      list_weight[p] <<- list_weight[p] + (list_weight[NM-(p-1)]/p)
      list_weight[NM-(p-1)] <<- list_weight[NM-(p-1)] - (list_weight[NM-(p-1)]/p)
      }
    
  }
}

weights(NM)
list_weight
```

################Definition of final weights################

``` {r Final weights}

#We have to check if several matrices have the same NA number.
#In this case, these matrices will have the same weight value.

sum = 0
final_w = 0

eq_weights <- function(i,j){
  for (c in 0:((i*j)-1)){
    eq <<- c(which((c == order_df$NumberNA) == TRUE))
    if (length(eq)>1){
      sum=0
      final_w=0
      for (p in eq){
        sum = sum + list_weight[p]
      }
      final_w = sum / length(eq)
      for (fw in eq) {
        list_weight[fw] <<- final_w
      }
    }
  }
}
eq_weights(i,j)
list_weight
```

################Addition of weights in the matrix################

``` {r Data frame}

#We add the column with weight values corresponding to each matrix. 

data = cbind(order_df, "Weight" = list_weight)
data
```

################Calculation of Likelihood and Weighted Likelihood################

``` {r Final computation}

#In this final function, we compute the weighted likelihood and the likelihood according to the functions we mentioned at the beginning.

WL = 0
L = 0
final_wlikelihood <- function(df2,NM){
  for (i in 1:NM){
    WL <<- WL + df2$Likelihood[i]*df2$Weight[i]
  }
  return(WL)
}
final_likelihood <- function(df2,NM){
  for (i in 1:NM){
    L <<- L + df2$Likelihood[i]
  }
  return(L)
}

final_wlikelihood(data,NM)
final_likelihood(data,NM)

```

################Retrieving data for comparison################

```{r Function}
initialization <- function(){
  NM<<-10                   #Number of matrices (max for time = 150)
  i<<-60                    #Lines in each matrix
  j<<-4                     #Columns in each matrix
  list_likelihood = c()   #Contains all likelihood values for all matrices
  list_NA = c()           #Contains the number of NA in each matrix
  list_weight = c()       #Contains all weight values for all matrices
  list_alpha = c()        #Contains all alpha values for all matrices
  list_beta = c()         #Contains all beta values for all matrices
  list_FP = c()           #Contains the number of FP in each matrix
  list_FN = c()           #Contains the number of FN in each matrix
  df1 <<- data.frame(matrix(ncol = 0, nrow = 0))
  order_df <<- data.frame(matrix(ncol = 0, nrow = 0))
  results_df <<- data.frame(matrix(ncol = 0, nrow = 0))
  WL <<- 0
  L <<- 0
}
 
final <- function(d){
  initialization()
  lists(NM,i,j)
  loglikelihood(NM,i,j,list_alpha, list_beta, list_FP, list_FN)
  while((Inf %in% list_likelihood)== TRUE) {
  lists(NM,i,j)
  loglikelihood(NM,i,j,list_alpha, list_beta, list_FP, list_FN)}
  NA_order()
  global_weights(NM)
  weights(NM)
  eq_weights(i,j)
  order_df <<- cbind(order_df, "Weight" = list_weight)
  DPWL[d] <<- final_wlikelihood(order_df,NM)
  DPL[d] <<- final_likelihood(order_df,NM)
}

```

################Plot################

```{r PLOT}
data_plot <- function(x) {
  DPWL <<- c()
  DPL <<- c()
  l <<- x
  for (i in 0:x){
    final(i)
  }
  results_df <<- data.frame("Likelihood" = DPL , "WLikelihood" = DPWL)
}

data_plot(NM)
results_df

plot(DPL, col="darkgreen",xlab = "Nth result of the function final()" )
plot(DPWL, col="red", xlab = "Nth result of the function final()")

matplot(1:l,cbind(DPL,DPWL),pch=20, col=c("darkgreen","red"), xlab = "Nth result of the function final()", ylab= "Values of DPL and DPWL")
legend("topleft", legend=c("DPL", "DPWL"),
       col=c("darkgreen", "red"), pch=20)
```

################Conclusion################

To compare the values obtained for likelihood and weighted likelihood I used the negative log which allows us to have more interpretable values.
We obtain these graphs and we can see that the likelihood values are always higher than the weighted one.
This allows us to conclude in several ways: 
  - This difference can be explained by the inclusion of NA. Indeed, our objective is to maximize the likelihood function. However, when a NA is present, we set 1. This is not without consequence because it increases the value of the likelihood when this should not be the case.
We can therefore conclude here that, although the values obtained by integrating the weights give us smaller values, these would therefore better reflect the results provided by the model.
  - The other conclusion is that there may be an error in the code and in the formulation of a formula, which could explain this big difference.
  - We can also identify a limit, which is to not take into account the matrix with the highest number of NA. This could also mislead the final result, decreasing it